---
title: "Dissertation_data_cleaning"
author: "Steven Bright"
date: "05/02/2022"
output: html_document
---

```{r Clean global environment}
remove(list=ls())
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Install packages}
#install.packages("psych")
#install.packages("tidyverse")
```

```{r Load packages}
library(psych)
library(tidyverse)
```

```{r Read in path script}
source("../Dissertation_paths.r")
```

#Read in data

```{r Read in data}
diss_data <- read_csv(paste0(data_path, data_name))
```

```{r Check data}
diss_data %>% 
  names()

diss_data %>%
  head()
```

#Formating changes

```{r Drop unwanted columns}
unwanted_column_names <- c(
  "StartDate",
  "EndDate",
  "Status",
  "IPAddress",
  "Progress",
  "Duration (in seconds)",
  "Finished",
  "RecordedDate",
  "ResponseId",
  "RecipientLastName",
  "RecipientFirstName",
  "RecipientEmail",
  "ExternalReference",
  "LocationLatitude",
  "LocationLongitude",
  "DistributionChannel",
  "UserLanguage",
  "id"
)

diss_data <- diss_data %>% 
  select(-all_of(unwanted_column_names))

diss_data %>% 
  names()
```

```{r Remove first two rows of data}
diss_data <- diss_data %>% 
  slice(c(-1, -2))
```

```{r Add participant_id column}
diss_data <- diss_data %>% 
  mutate(ID = 1:n()) %>% 
  select(ID, everything())

diss_data %>% 
  head()
```

```{r Rename columns}

#Put correct column names into a vector
correct_column_names <- c(
  "ID",
  "Participant_consent",
  "Dem.age",
  "Dem.sex",
  "Dem.ethnic_origin",
  "Wellbeing.optimistic_future",
  "Wellbeing.feeling_useful",
  "Wellbeing.feeling_relaxed",
  "Wellbeing.interested_others",
  "Wellbeing.spare_energy",
  "Wellbeing.dealing_with_problems",
  "Wellbeing.thinking_clearly",
  "Wellbeing.feeling_good_self",
  "Wellbeing.feeling_close_others",
  "Wellbeing.feeling_confident",
  "Wellbeing.makeup_mind",
  "Wellbeing.feeling_loved",
  "Wellbeing.interested_new_things",
  "Wellbeing.feeling_cheerful",
  "Selfesteem.satisfied_with_self",
  "Selfesteem.no_good_at_times",
  "Selfesteem.good_qualities",
  "Selfesteem.well_as_most_people",
  "Selfesteem.not_much_proud",
  "Selfesteem.feel_useless",
  "Selfesteem.equal_worth_others",
  "Selfesteem.wish_more_respect",
  "Selfesteem.feel_faliure",
  "Selfesteem.positive_attitude",
  "DERS.clear_about_feelings.1",
  "DERS.pay_attention_feelings.2",
  "DERS.emotions_overwhelming.3",
  "DERS.no_idea_feelings.4",
  "DERS.difficulty_sense_feelings.5",
  "DERS.attentive_to_feelings.6",
  "DERS.know_exactly_feelings.7",
  "DERS.care_about_feelings.8",
  "DERS.confused_what_i_feel.9",
  "DERS.acknowledge_emotions.10",
  "DERS.angry_fo_feeling_upset.11",
  "DERS.embarrassed_feeling_upset.12",
  "DERS.difficulty_working_upset.13",
  "DERS.out_of_control_upset.14",
  "DERS.remain_long_time_upset.15",
  "DERS.end_up_depressed_upset.16",
  "DERS.feelings_valid_upset.17",
  "DERS.difficulty_focusing_upset.18",
  "DERS.feel_out_of_control_upset.19",
  "DERS.get_things_done_upset.20",
  "DERS.ashamed_feeling_upset.21",
  "DERS.know_feel_better.22",
  "DERS.feel_weak_upset.23",
  "DERS.remain_control_upset.24",
  "DERS.guilty_feeling_upset.25",
  "DERS.difficulty_concentrating_upset.26",
  "DERS.difficulty_controlling_behaniours_upset.27",
  "DERS.nothing_make_feel_better_upset.28",
  "DERS.irritated_feeling_upset.29",
  "DERS.feel_bad_upset.30",
  "DERS.wallow_upset.31",
  "DERS.lose_control_upset.32",
  "DERS.difficulty_thinking_else_upset.33",
  "DERS.figure_out_feelings_upset.34",
  "DERS.long_time_feel_better_upset.35",
  "DERS.emotions_overwhelming_upset.36"
)

#Change column names
names(diss_data) <- correct_column_names

diss_data %>% 
  colnames()

diss_data %>% 
  head()
```

#Data cleaning

```{r Change column data type to numeric}
diss_data <- diss_data %>% 
  map_df(as.numeric)
```


Participants who didn't reach the end of the survey should be excluded 
because we indicated that they could close the broswer to withdraw.

I have identified the partial responders by viewing the tibble and checking
which participants consistently left NA responses before the end of the survey.

```{r Remove partial responders}
#diss_data %>% 
#  View()

#Remove the partial responders
diss_data <- diss_data %>% 
  slice(-c(19, 20, 21, 22, 23, 24, 66, 67, 90, 149, 166))
```

#Create total sum scores

There are three sum scores to create:
1) Mental wellbeing (0-70)
2) Self-esteem (10-40)
3) DERS (0-180)

```{r Create mental wellbeing sum score}
#Sum all the wellbeing scores
diss_data$Wellbeing.sum_score <- diss_data %>% 
  select(starts_with("Wellbeing.")) %>% 
  rowSums(., na.rm = TRUE)

#Check values
diss_data %>% 
  select(Wellbeing.sum_score) %>% 
  summary()
```

```{r Create self-esteem sum score}
#Sum all the self-esteem scores
diss_data$Selfesteem.sum_score <- diss_data %>% 
  select(starts_with("Selfesteem.")) %>% 
  rowSums(., na.rm = TRUE)

#Check values
diss_data %>% 
  select(Selfesteem.sum_score) %>% 
  summary()
```

```{r Create keys variable for DERS}
#1 for normal coding, -1 for reverse coding
ders_sum_score_keys <- rep(1, times = 36)
```

```{r Get names of ders variables}
ders_variable_names <- diss_data %>% 
  select(starts_with("DERS.")) %>% 
  colnames()
```

```{r Create ders total sum score}
# calculate total/sum score
ders_total_score <- scoreItems(keys = ders_sum_score_keys,
                               items = diss_data[ders_variable_names],
                               totals = TRUE, #total sum score
                               missing = TRUE, #will include individuals that have missing values
                               impute = 'none',
                               min = 1, 
                               max = 5
                               )

#Add in total score column to data frame
diss_data <- diss_data %>%
  mutate(DERS.sum_score = 
           as.numeric(ders_total_score$scores))

#check the values
diss_data %>%
  select(DERS.sum_score) %>% 
  summary(ders.sum_score)
```

##Create DERS subscales

Unlike the MW and SE scales, the DERS is a multidimensional scale with a total of 6 subscales. The subscales include:

Note: "R" stands for "Reverse-coded".

1) Non-acceptance (Items 11, 12, 21, 23, 25, 29)
2) Difficulties engaging in goal-directed behaviour (Items 13, 18, 20R, 26, 33)
3) Impulse control difficulties (Items 3, 14, 19, 24R, 27, 32)
4) Lack of emotional awareness (Items 2R, 6R, 8R, 10R, 17R, 34R)
5) Limited access to emotion regulation strategies (Items 15, 16, 22R, 28, 30, 31, 35, 36)
6) Lack of emotional clarity (Items 1R, 4, 5, 7R, 9)

### 1) Non-acceptance subscale sum score

```{r Create the non-acceptance sum score}
#Get names of DERS non-acceptance items
DERS_non_acceptance_names <- diss_data %>% 
  select(DERS.angry_fo_feeling_upset.11,
  DERS.embarrassed_feeling_upset.12,
  DERS.ashamed_feeling_upset.21,
  DERS.feel_weak_upset.23,
  DERS.guilty_feeling_upset.25,
  DERS.irritated_feeling_upset.29
  ) %>% 
  colnames()

#Sum all the non-acceptance items
diss_data$DERS.non_acceptance_sum_score <- diss_data %>% 
  select(all_of(DERS_non_acceptance_names)) %>% 
  rowSums(., na.rm = TRUE)

#Check values
diss_data %>% 
  select(DERS.non_acceptance_sum_score) %>% 
  summary()
```

### 2) Difficulties in engaging in goal-directed behaviour

```{r Create keys for DERS goals subscale}
#1 for normal coding, -1 for reverse coding
ders_goals_keys <- c(
  1,
  1,
  -1,
  1,
  1
)
```

```{r Get names of DERS goals variables}
ders_goals_variable_names <- diss_data %>% 
  select(DERS.difficulty_working_upset.13,
  DERS.difficulty_focusing_upset.18,
  DERS.get_things_done_upset.20,
  DERS.difficulty_concentrating_upset.26,
  DERS.difficulty_thinking_else_upset.33
  ) %>% 
  colnames()
```

```{r Create ders goals sum score}
# calculate total/sum score
ders_goals_score <- scoreItems(keys = ders_goals_keys,
                               items = diss_data[ders_goals_variable_names],
                               totals = TRUE, #total sum score
                               missing = TRUE, #will include individuals that have missing values
                               impute = 'none',
                               min = 1, 
                               max = 5
                               )

#Add in total score column to data frame
diss_data <- diss_data %>%
  mutate(DERS.goals_sum_score = 
           as.numeric(ders_goals_score$scores))

#Check the values
diss_data %>%
  select(DERS.goals_sum_score) %>% 
  summary(DERS.goals_sum_score)
```

### 3) Impulse control difficulties

```{r Create ders impulse keys}
#1 for normal coding, -1 for reverse coding
ders_impulse_keys <- c(
  1,
  1,
  1,
  -1,
  1,
  1
)
```

```{r Get names of DERS impulse variables}
ders_impulse_variable_names <- diss_data %>% 
  select(DERS.emotions_overwhelming.3,
         DERS.out_of_control_upset.14,
         DERS.feel_out_of_control_upset.19,
         DERS.remain_control_upset.24,
         DERS.difficulty_controlling_behaniours_upset.27,
         DERS.lose_control_upset.32
         ) %>% 
  colnames()
```

```{r Create ders impulse sum score}
# calculate total/sum score
ders_impulse_score <- scoreItems(keys = ders_impulse_keys,
                               items = diss_data[ders_impulse_variable_names],
                               totals = TRUE, #total sum score
                               missing = TRUE, #will include individuals that have missing values
                               impute = 'none',
                               min = 1, 
                               max = 5
                               )

#Add in total score column to data frame
diss_data <- diss_data %>%
  mutate(DERS.impulse_sum_score = 
           as.numeric(ders_impulse_score$scores))

#Check the values
diss_data %>%
  select(DERS.impulse_sum_score) %>% 
  summary(DERS.impulse_sum_score)
```

### 4) Lack of emotional awareness

```{r Create ders awareness keys}
#1 for normal coding, -1 for reverse coding
ders_awareness_keys <- rep(-1, times = 6)
```

```{r Get names of DERS awareness variables}
ders_awareness_variable_names <- diss_data %>% 
  select(DERS.pay_attention_feelings.2,
         DERS.attentive_to_feelings.6,
         DERS.care_about_feelings.8,
         DERS.acknowledge_emotions.10,
         DERS.feelings_valid_upset.17,
         DERS.figure_out_feelings_upset.34
         ) %>% 
  colnames()
```

```{r Create ders awareness sum score}
# calculate total/sum score
ders_awareness_score <- scoreItems(keys = ders_awareness_keys,
                               items = diss_data[ders_awareness_variable_names],
                               totals = TRUE, #total sum score
                               missing = TRUE, #will include individuals that have missing values
                               impute = 'none',
                               min = 1, 
                               max = 5
                               )

#Add in total score column to data frame
diss_data <- diss_data %>%
  mutate(DERS.awareness_sum_score = 
           as.numeric(ders_awareness_score$scores))

#Check the values
diss_data %>%
  select(DERS.awareness_sum_score) %>% 
  summary(DERS.awareness_sum_score)
```

### 5) Limited access to strategies

```{r Create ders srategies keys}
#1 for normal coding, -1 for reverse coding
ders_strategies_keys <- c(
  1,
  1,
  -1,
  1,
  1,
  1,
  1,
  1
)
```

```{r Get names of DERS strategies variables}
ders_strategies_variable_names <- diss_data %>% 
  select(DERS.remain_long_time_upset.15,
         DERS.end_up_depressed_upset.16,
         DERS.know_feel_better.22,
         DERS.nothing_make_feel_better_upset.28,
         DERS.feel_bad_upset.30,
         DERS.wallow_upset.31,
         DERS.long_time_feel_better_upset.35,
         DERS.emotions_overwhelming_upset.36
         ) %>% 
  colnames()
```

```{r Create ders strategies sum score}
# calculate total/sum score
ders_strategies_score <- scoreItems(keys = ders_strategies_keys,
                               items = diss_data[ders_strategies_variable_names],
                               totals = TRUE, #total sum score
                               missing = TRUE, #will include individuals that have missing values
                               impute = 'none',
                               min = 1, 
                               max = 5
                               )

#Add in total score column to data frame
diss_data <- diss_data %>%
  mutate(DERS.strategies_sum_score = 
           as.numeric(ders_strategies_score$scores))

#Check the values
diss_data %>%
  select(DERS.strategies_sum_score) %>% 
  summary(DERS.strategies_sum_score)
```

### 6) Lack of emotional clarity

```{r Create ders clarity keys}
#1 for normal coding, -1 for reverse coding
ders_clarity_keys <- c(
  -1,
  1,
  1,
  -1,
  1
)
```

```{r Get names of DERS clarity variables}
ders_clarity_variable_names <- diss_data %>% 
  select(DERS.clear_about_feelings.1,
         DERS.no_idea_feelings.4,
         DERS.difficulty_sense_feelings.5,
         DERS.know_exactly_feelings.7,
         DERS.confused_what_i_feel.9
         ) %>% 
  colnames()
```

```{r Create ders clarity sum score}
# calculate total/sum score
ders_clarity_score <- scoreItems(keys = ders_clarity_keys,
                               items = diss_data[ders_clarity_variable_names],
                               totals = TRUE, #total sum score
                               missing = TRUE, #will include individuals that have missing values
                               impute = 'none',
                               min = 1, 
                               max = 5
                               )

#Add in total score column to data frame
diss_data <- diss_data %>%
  mutate(DERS.clarity_sum_score = 
           as.numeric(ders_clarity_score$scores))

#Check the values
diss_data %>%
  select(DERS.clarity_sum_score) %>% 
  summary(DERS.clarity_sum_score)
```

#Write clean data

```{r Re-order columns}
diss_data %>%
  colnames()

diss_data <- diss_data %>% 
  select(ID, 
         Participant_consent,
         Dem.age,
         Dem.sex,
         Dem.ethnic_origin,
         Wellbeing.sum_score,
         Selfesteem.sum_score,
         DERS.sum_score,
         DERS.non_acceptance_sum_score,
         DERS.goals_sum_score,
         DERS.impulse_sum_score,
         DERS.awareness_sum_score,
         DERS.strategies_sum_score,
         DERS.clarity_sum_score,
         everything())
```

```{r Write clean data to be used in SPSS}
write.csv(diss_data,
          file = paste0(clean_data_path,
                        "clean_diss_data.csv"))
```
